### 2025-11-16 thesis/theoretical-foundations/llms.tex
Extend the llm theory

Large language models (LLMs) represented a paradigm shift in natural language
processing, demonstrating state of the art capabilities in understanding, reasoning and
text generation. These models form the core of most modern NL2SQL systems
like \textsc{XiYan-SQL}, \textsc{DAIL} and others. They have a wide language understanding as they have been trained on large text corpora,
are exposed to different domains of traning data and have an inherent
understanding of the SQL syntax as they have been exposed to it during
pre-training. This shifted the approach while developing NL2SQL systems to
focus on in-context learning methods instead of training dedicated models from
scratch.

\paragraph{Feedforward and Residual Connections}

@@

\paragraph{Positional Encoding}

@@

Large language models are trained in a two-stage process that separates the
training of general language understanding from from task-specific
capabilities.

@@

Fine-tuning is the process of adapting a pre-trained model, for example for
solving NL2SQL tasks, by continuing model training on a labeled dataset of
question-answer pairs (eg, natural language to sql). During fine-tuning, the
model can learn to:

\begin{itemize}
    \item Map natural language to database schema elements (also known as schema-linking)
    \item Generate SQL queries following proper grammar
    \item Compose more complex SQL queries involving joins, aggregations, and subqueries
    \item Understand domain-specific terminology (depending on the dataset)
\end{itemize}

@@

>>>>

Large language models (LLMs) are a recent advancement built on top of the
transformer architecture for natural language processing, demonstrating state
of the art capabilities in understanding, reasoning and text generation. These
models form the core of most modern NL2SQL systems like \textsc{XiYan-SQL},
\textsc{DAIL} and others. They have a wide language understanding as they have
been trained on large text corpora, are exposed to different domains of traning
data and have an inherent understanding of the SQL syntax as they have been
exposed to it during pre-training. This shifted the approach while developing
NL2SQL systems to focus on in-context learning methods instead of training
dedicated models from scratch.

\subsubsection{Feedforward and Residual Connections}

@@

\subsubsection{Positional Encoding}

@@

Large language models are trained in a two-fold process that separates the
training of general language understanding from from task-specific
capabilities.

@@

Fine-tuning is the process of adapting a pre-trained model, for example for
solving NL2SQL tasks, by continuing model training on a labeled dataset
(eg, natural language question to expected sql query). During fine-tuning, the
model can learn to map natural language questions to expected SQL queries when
being presented with a question and a database schema.

@@

\subsubsection{Prompt Engineering}

Prompt engineering is the process of structuring the input to LLMs in a way
where they adhere to desired behaviours. For NL2SQL systems previous research
has found diverging effectiveness between different example, schema and
question presentation mechanisms \citep{DAIL, OmniSQL}.

\subsubsection{Chain-of-Thought Reasoning}

Prompting or training models to output their chain-of-thought encourages models 
to generate reasoning traces before producing their final answer to a question
or input. For NL2SQL, this usually involves generating explanations of query
logic, analysing the input database schema, and planning the query structure
before generating a query.

Fine-tuned models like \textsc{OmniSQL} are fine-tuned to produce
chain-of-thought output during generation in markdown which enables debugging
model behaviour and understanding the query generation approach.

\subsubsection{Model Limitations}

Understanding the limitations of large language models is essential for
designing robust NL2SQL systems.

\paragraph{Hallucination}

As LLMs are just predicting a sequence of output tokens, they have no inherent
understanding of the actual problem domain and solution space they are
presented with. Thus in NL2SQL systems, LLMs might generate invalid SQL queries
which might reference non existent tables, columns or relationships, output
invalid syntax or generate promising but semantically invalid queries.

These flaws are inherent to a text based representation and are non
recoverable, although through candidate validation and self-refinement
approaches most obviously invalid queries can be recovered. Detecting semantic
invalidity is a hard research problem on its own.

\paragraph{Context Window Limitations}

As the underlying transformer architecture is still bound to finite context
windows, extensive database schemas or long lists of examples can exceed the
window, thus causing the query generation to fail.
